---
title: "vignette-cutouts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette-cutouts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PaluckMetaSOP)
```

This vignette explains and demonstrates the `d_calc.R` and `var_d_calc.R` functions.

`d_calc` is a calculator that converts a study's results into an estimate of a standardized mean effect, typically called `d` or ∆ (Delta); `var_d_calc` provides estimates of the variance of ∆. These variables are the foundation of a meta-analysis.

This vignette is not an exhaustive overview of why, how and when to convert a study into a single point estimate -- see Cooper, Hedges, and Valentine for that. Instead, we show you how we've used these in the past to make meta-analyses go more smoothly.

Most of our examples will be from [The Contact Hypothesis Re-evaluated](https://doi.org/10.1017/bpp.2018.25), with a few examples drawn from [Preventing Sexual Violence —A Behavioral Problem Without a Behaviorally-Informed Solution](https://osf.io/preprints/psyarxiv/xgbzj).

This vignette first defines the functions' necessary inputs; then works through examples of each type of statistical information that `d_calc` is designed to interpret; then provides two examples of how to incorporate these functions into your workflow.

### The core parameters:

`d_calc` takes in between 2-5 inputs, depending on the value of the first input: \* `stat_type`, the type of statistical information that a study's results are reported as; `stat`, the value of that stat_type (e.g. F = 1.0 or reg_coef (regression coefficient) = 0.2); \* `sample_sd`, which is an estimate of the standard deviation of the dependent variable, drawn either from the entire sample (for cohen's d) or from the control group (for Glass's ∆). This parameter will not be necessary if the stat_type already conveys information about the variance (F-test, t-test, or `d`); `n_t`, the sample size for the treatment group; and `n_c`, the sample size for the control group.

`d_calc` then returns a single number, which corresponds to Cohen's `d` or Glass's ∆ depending on what `sample_sd` you provided.

`var_d_calc` takes as input:

-   d, typically the output of `d_calc`;
-   `n_t`, the sample size for the treatment group;
-   `n_c`, the sample size for the control group.

We'll now work through each of the possible values of `stat_type`
